---
title: "effects"
author: "Rich"
date: "2024-01-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file takes the model output and plots effects from the posterior draws.

## load the libraries that we will be using ## 

## install ##

```{r install-pkg}
# install.packages(c("tidyverse", "RColorBrewer", "patchwork", "brms",
#                    "tidybayes", "bayesplot", "future", "sjPlot"))
```

take a snapshot of loaded packages and update the lock.file using renv

```{r snapshot-renv}
# take a snapshot and update the lock.file
# renv::snapshot() # this is only necessary when new packages or installed or packages are updated.
```

## load ##

```{r load-pkg}
pkg <- c("tidyverse", "RColorBrewer", "patchwork", "brms", 
         "tidybayes", "bayesplot", "future", "parallel", "sjPlot")

lapply(pkg, library, character.only = TRUE)
```


# Read in previously saved brms model object files ##

If available, this saves time computing them each time 

```{r}
b1 <- readRDS("models/b1.rds")
```

## Look at the point and interval estimates for model bpp5.3 (the full model) ##

using posterior_samples and tidybayes

```{r}
post <- as_draws_df(b1) %>%  #
  select(-lp__) %>% 
  as_tibble()

post_summary <- posterior_summary(b1, robust = TRUE)

# all fixed effects
post_qi_b <- post %>%
  select(starts_with("b_")) %>% 
  pivot_longer(everything()) %>% 
  group_by(name) %>% 
  median_qi(value)
head(post_qi_b)

# varying effects - SD
post_qi_sd <- post %>%
  select(starts_with("sd_")) %>% 
  pivot_longer(everything()) %>% 
  group_by(name) %>% 
  median_qi(value)
head(post_qi_sd)

# participant variation
vary_vals_pt <- post %>%
  select(starts_with("r_")) %>% 
  pivot_longer(everything()) %>% 
  group_by(name) %>% 
  filter(str_detect(name, "pid")) %>% 
  median_qi(value)
head(vary_vals_pt)
```

## Visualise fixed effects ##

Using posterior_samples and tidybayes

```{r}
## fixed effects of interest, excluding basic intercepts for the ordinal model 
tidy_fixed <- post %>% 
  select(starts_with("b_")) %>% # select and rename in simpler labels
  pivot_longer(everything()) %>% # move from wide to long
  mutate(key = factor(name, levels=unique(name)),
         term = if_else(str_detect(key, "Intercept"), "none",
                if_else(str_detect(key, "left"), "left", "right")),
         term = factor(term, levels=c("left", "none", "right")))  
head(tidy_fixed)
tail(tidy_fixed)

check.labels <- tidy_fixed %>% 
 group_by(key, term) %>% 
 distinct(key, term)
check.labels

# remove the intercept, as it is on a different scale
tidy_fixed <- tidy_fixed %>% 
  filter(term != "none")
head(tidy_fixed)
str(tidy_fixed)

# plot
p_tidy_fixed <- ggplot(tidy_fixed, aes(x = value, y = fct_rev(term), fill=term)) +  
  geom_vline(xintercept = 0, color = "grey", alpha = 5/10) +
  stat_halfeye(point_size = 1) +
  labs(title = "Coefficient plot for fixed effects (predictors)",
       x = NULL, y = NULL) +
  theme_bw() +
  scale_fill_brewer(palette="Dark2") +
  theme(panel.grid   = element_blank(),
        panel.grid.major.y = element_line(color = alpha("firebrick4", 1/2), linetype = 3),
        axis.text.y  = element_text(hjust = 0),
        axis.ticks.y = element_blank(),
        legend.position = "none")
p_tidy_fixed

ggsave ("figures/b1_fixef.jpeg")
```

# make a table of fixed effects #

```{r}
fix_tab <- post_qi_b %>%
  mutate(name = factor(name, levels=unique(name))) %>%
  rename(term = name) %>% 
  select(term, value, .lower, .upper)
fix_tab

tab_df(fix_tab,
       alternate.rows = T,
       title="fixed effects",
       file="tables/fixed.doc")
```

# Visualise varying effects #

using posterior samples and tidybayes.

# Variance at the group level (i.e., intercept for item and intercepts and effects across participants)

```{r}
## varying effects of interest, including item, as well as participant and effects across pts
tidy_vary <- post %>% 
  select(contains("sd_")) %>% 
  pivot_longer(everything()) %>% 
  mutate(key = factor(name, levels=unique(name)),
         term = if_else(str_detect(key, "Intercept"), "none",
                if_else(str_detect(key, "left"), "left", "right")),
         term = factor(term, levels=c("left", "none", "right")))  
head(tidy_vary)
tail(tidy_vary)

check.vary <- tidy_vary %>% 
  group_by(key, term) %>% 
  distinct(key, term)
check.vary

p_tidy_vary <- ggplot(tidy_vary, aes(x = value, y = fct_rev(term), fill=term)) +  
  geom_vline(xintercept = 0, color = "grey", alpha = 5/10) +
  stat_halfeye() +
  labs(title = "SD plot for varying effects",
       x = NULL, y = NULL) +
  theme_bw() +
  scale_fill_brewer(palette="Dark2") +
  theme(panel.grid   = element_blank(),
        panel.grid.major.y = element_line(color = alpha("firebrick4", 1/2), linetype = 3),
        # axis.text.y  = element_blank(),
        # axis.ticks.y = element_blank(),
        legend.position = "none") 
  # coord_cartesian(xlim =c(0, 1.5)) +
  # scale_x_continuous(breaks=seq(0,1.5,0.5))
p_tidy_vary

ggsave ("figures/varypp.jpeg")

```

# Variance within group levels i.e., at the individual item and participant level #

Note - this should be in a centred metric, so that deviation from the group average (zero) can be visualised.

# filter and wrangle varying effects within participants #

# participant level #

```{r}
# 12 pts x 3 terms = 36
vary_pt <- post_summary %>%
  data.frame() %>% 
  rownames_to_column("param") %>% 
  filter(str_detect(param, "r_pid"),
         str_detect(param, "cor", negate = TRUE)) %>% 
  mutate(term = factor(param, levels=unique(param)),
         term = if_else(str_detect(term, "Intercept"), "none",
                if_else(str_detect(term, "left"), "left", "right")),
         term = factor(term, levels=c("left", "none", "right"))) %>% 
  arrange(term, Estimate) %>% 
  mutate(rank = rep(1:12, times=3),
         outlier = if_else(Q2.5 > 0 | Q97.5 < 0, "yes", "no"), 
         outlier = factor(outlier, levels = c("no", "yes")))
head(vary_pt)

check.vary.pt <- vary_pt %>% 
  distinct(term)
check.vary.pt
```

# plot varying effects via caterpillar plots #

```{r}
p_tidy_varyvals <- ggplot(vary_pt, aes(x=rank, y=Estimate, colour=outlier))+
  geom_pointrange(aes(ymin=Q2.5,ymax=Q97.5),fatten = 0.5)+
  geom_hline(yintercept=0, colour = "red")+
  theme_bw()+
  scale_colour_manual(values=c("darkgrey","black")) +
  labs(title = "Varying effects by participant",
       x = "participant")+
  # theme(axis.text.x  = element_blank())+
  guides(colour = "none") +
  facet_wrap(~term) +
  scale_x_continuous(breaks = 1:12)
p_tidy_varyvals

ggsave ("figures/varyvalspp.jpeg")
```