---
title: "model"
author: "Rich"
date: "2024-01-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file fits Bayesian multi-level regression models for Exp2a of Holmes et al., 2023.

## load the libraries that we will be using ## 

## install ##

```{r install-pkg}
# install.packages("remotes")
# remotes::install_github("stan-dev/cmdstanr")
# 
# install.packages("devtools")
# devtools::install_github("jmgirard/standist")
# 
# install.packages(c("tidyverse", "RColorBrewer", "patchwork", "brms",
#                     "tidybayes", "bayesplot", "future", "modelr"))
```

take a snapshot of loaded packages and update the lock.file using renv

```{r snapshot-renv}
# take a snapshot and update the lock.file
# renv::snapshot() # this is only necessary when new packages or installed or packages are updated.
```

## load ##

```{r load-pkg}
pkg <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel", "modelr")

lapply(pkg, library, character.only = TRUE)
```

## settings ##

```{r set-options}
options(brms.backend = "cmdstanr",
        mc.cores = parallel::detectCores(),
        future.fork.enable = TRUE,
        future.rng.onMisuse = "ignore") ## automatically set in RStudio

supportsMulticore()

detectCores()
```

## plot settings ##

```{r}
## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

theme settings for ggplot

```{r, eval = F}
theme_set(
  theme_bw() +
    theme(text = element_text(size = 18), 
          title = element_text(size = 18),
          legend.position = "bottom")
)
```

## section 1 - read in data and create factors, read in models if computed ##

```{r}
data <- read_csv("data/data.csv")
head(data)
```

create factors for modelling purposes

a couple of options:
- no stimulation seems like a natural reference category
- anova style deviation coding also seems sensible.

let's start with reference coding and go from there.

```{r}
data <- data %>% 
  mutate(condition = factor(condition, 
                            levels = c("none", "left", "right")))
head(data)
```

read in previous models

```{r}
b1 <- readRDS("models/b1.rds")
summary(b1)
```

## section 2 - build some models ##

## b0 - intercepts only ##

this is just to get things going i.e., a model without any predictors.

# formula #

```{r}
formula = bf(count ~ 1)
```

# check the priors available #

we'll just go with a gaussian() to keep things simple

```{r}
get_prior(formula,
          data = data, family = gaussian())
```

## visualise priors ##

here we would normally visualise priors of interest to make a judgment about what
would constitute weakly informative priors. 

https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations

```{r}
visualize("normal(14, 1)", "normal(14, 2)", "normal(14, 3)",
          xlim = c(5, 25))
```

## set priors ##

```{r}
priors <- c(
  set_prior("normal(14, 3)", class = "Intercept"),
  set_prior("normal(0, 1)", class = "sigma")
)
```

# run the model #

```{r}
plan(multicore)
b0 <- brm(formula = formula,
        data = data, family = gaussian(),
        prior = priors,
        iter = 2000, warmup = 1000, cores = 8, chains = 4,
        save_pars = save_pars(all=TRUE),
        seed = 123,
        file = "models/b0")
summary(b0)
```

## take a look ##

chains

```{r}
plot(b0)
```

pp check

```{r}
pp_b0 <- pp_check(b0, ndraws = 100)
pp_b0
```


## b1 - add a predictor for condition plus varying intercepts and effects by pid  ##

# formula #

```{r}
formula = bf(count ~ 1 + condition +
               (1 + condition | pid))
```

# check the priors available #

we'll just go with a gaussian() to keep things simple

```{r}
get_prior(formula,
          data = data, family = gaussian())
```

## visualise priors ##

here we would normally visualise priors of interest to make a judgment about what
would constitute weakly informative priors. 

https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations

```{r}
visualize("normal(0, 1)", "normal(0, 2)", "normal(0, 3)",
          xlim = c(-4, 4))
```

## set priors ##

```{r}
priors <- c(
  set_prior("normal(14, 3)", class = "Intercept"),
  set_prior("normal(0, 1)", class = "b"),
  set_prior("normal(0, 1)", class = "sd"),
  set_prior("normal(0, 1)", class = "sigma"),
  set_prior("lkj(2)", class = "cor")
)

```

# run the model #

```{r}
plan(multicore)
b1 <- brm(formula = formula,
        data = data, family = gaussian(),
        prior = priors,
        iter = 2000, warmup = 1000, cores = 8, chains = 4,
        save_pars = save_pars(all=TRUE),
        seed = 123,
        file = "models/b1")
summary(b1)
```

## take a look ##

chains

```{r}
plot(b1)
```

pp check

```{r}
pp_b1 <- pp_check(b1, type = "dens_overlay_grouped", group = "condition", 
                  ndraws = 100)
pp_b1
```


We would probably centre the dv for the real thing.

And we may also run a model with condition coded as a deviation coded variable.
