---
title: "sims"
author: "Rich"
date: "2024-01-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file simulates data and runs power analyses based on the modelling of
Exp2a from Holmes et al., 2023.

This is a little redundant in some ways, as with simple designs, then G*Power
could a quick and dirty job, if you plug in a standardised effect size between
two conditions, for example. On that note, Cohen's dz is over 1 for key comparisons,
which means N=12 would give you > 90% power.

But that kind of quick and dirty approach ignores the multi-level structure to the
data and varying effects by participant, which is the benefit of the multi-level
modelling approach taken here.


Most of the below is based on two sources:

1) The {faux} package: https://debruine.github.io/faux/

2) Solomon Kurz's blog post: https://solomonkurz.netlify.app/blog/bayesian-power-analysis-part-i/

## load the libraries that we will be using ## 

## install ##

you may only need the additional faux package here

```{r}
# install.packages(c("faux"))
```

if not, then run the next chunk, which installs them all.

```{r install-pkg}
# install.packages("remotes")
# remotes::install_github("stan-dev/cmdstanr")
# 
# install.packages("devtools")
# devtools::install_github("jmgirard/standist")
# 
# install.packages(c("tidyverse", "RColorBrewer", "patchwork", "brms",
#                    "tidybayes", "bayesplot", "patchwork", "future", "faux"))
```

take a snapshot of loaded packages and update the lock.file using renv

```{r snapshot-renv}
# take a snapshot and update the lock.file
# renv::snapshot() # this is only necessary when new packages or installed or packages are updated.
```

## load ##

```{r load-pkg}
pkg <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel", "modelr",
         "faux")

lapply(pkg, library, character.only = TRUE)
```

## settings ##

```{r set-options}
options(brms.backend = "cmdstanr",
        mc.cores = parallel::detectCores(),
        future.fork.enable = TRUE,
        future.rng.onMisuse = "ignore") ## automatically set in RStudio

supportsMulticore()

detectCores()
```

## plot settings ##

```{r}
## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

theme settings for ggplot

```{r, eval = F}
theme_set(
  theme_bw() +
    theme(text = element_text(size = 18), 
          title = element_text(size = 18),
          legend.position = "bottom")
)
```

## section 1 - read in prior models and data ##

this is just useful as a reminder of key values and structures

data

```{r}
data <- read_csv("data/data.csv")
head(data)
```

model fit

```{r}
b1 <- readRDS("models/b1.rds")
summary(b1)
```

model fit of simulated data

```{r}
# based on model b1
# fit <- readRDS("models/sims/fit.rds") 
# summary(fit)
```

## section 2 - simulate some data ##

## based on the count dv and b1 model ##

define some values

```{r}
# define parameters
# specify some features of the design
subj_n = 12  # number of subjects
rep_n = 8 # number of trial repeats 

# set fixed effects
b0 = 14.99    # intercept (no stimulation)
b1 = -1.71    # fixed effect of left stim < no stim
b2 = -0.16    # fixed effect of right stim < no stim

# set varying effects (by subject)
u0s_sd = 1.14   # varying intercept SD 
u1s_sd = 1.39   # varying b1 slope SD 
u2s_sd = 0.40   # varying b2 slope SD

# set correlations between varying effects
# cors between mean average effects first
r01s = -0.52   # correlation between varying effects 0 and 1 
r02s = -0.11   # correlation between varying effects 0 and 2 
r12s = 0.34  # correlation between varying effects 1 and 2 

# sigma
sigma_sd = 1.26 # error SD
```

make a correlation matrix 

```{r}
cors = c(r01s, r02s,
         r12s)
```

setup the data structure

```{r}
# make it reproducible
set.seed(1)

# set up data structure
d1 <- add_random(subj = subj_n, rep = rep_n) %>%
  # add and recode categorical variables
  add_within("subj", condition = c("none", "left", "right")) %>%
  add_contrast("condition", "treatment", add_cols = TRUE, 
               colnames = c("left", "right")) %>%
  # add random effects 
  add_ranef("subj", u0s = u0s_sd, u1s = u1s_sd, u2s = u2s_sd,
            .cors = cors) %>%
  add_ranef(sigma = sigma_sd) %>%
  # calculate dv
  mutate(count = b0 + u0s + (b1 + u1s) * left + (b2 + u2s) * right + sigma)  

head(d1)
str(d1)
summary(d1)

# save initial data
write_csv(d1, "data/sims/d1.csv") # 
```

data.check

```{r}
data.check <- d1 %>% 
  distinct(rep, condition)
data.check
```

density plot 

```{r}
p2.1 <- ggplot(d1, aes(x=count, fill=condition)) +
   geom_density(alpha = 0.3, colour = "darkgrey") +
   scale_fill_brewer(palette = "Dark2")+
   theme(panel.grid = element_blank()) +
   ggtitle("count by stimulation condition")
p2.1

ggsave ("figures/sims/d1_count_density.jpeg")
```

create some group average summary data

per pid

```{r}
summary_pid <- d1 %>% 
  group_by(subj, condition) %>% 
  summarise(mean = mean(count)) %>% 
  mutate(condition = factor(condition, 
                            levels = c("left", "none", "right")))
summary_pid
```

at the group level

```{r}
summary <- d1 %>% 
  group_by(condition) %>% 
  summarise(mean = mean(count),
            sd = sd(count),
            n = length(unique(subj)), # n here is the total subjs
            sem = (sd/sqrt(length(unique(subj)))),
            ci = sem*1.96) %>% 
  mutate(condition = factor(condition, 
                            levels = c("left", "none", "right")))
summary
```

violin

```{r}
p2.2 <- ggplot(summary_pid, aes(x=condition, y = mean, 
                                fill = condition, colour = condition)) +
   geom_jitter(position=position_jitterdodge(dodge.width =1), 
               alpha = 1, colour = "darkgrey") +
   geom_line(aes(group=subj), 
             colour = "darkgrey") +
   geom_violin(alpha = 0.7) +
   geom_point(data = summary, 
             aes(y = mean), size = 3, position=pd2, colour="black") +
   geom_errorbar(data = summary,
                aes(y = mean, ymin = mean-sem, ymax = mean +sem),
                width=.2, position=pd2, colour = "black") +
   geom_line(data = summary, 
            aes(y = mean, group=1),
            position=pd2,  colour = "black") +
   scale_fill_brewer(palette = "Dark2") +
   scale_colour_brewer(palette = "Dark2") +
   ggtitle("simdat1 count by stimulation condition")
p2.2

ggsave ("figures/sims/d1_count_violin.jpeg")
```

this also looks fairly sensible

## section 3 - build an initial model with one simulated dataset ##

This initial model will serve as the basis for other models to be built from using
the 'update' function in brms. It saves a lot of time this way.

The models will be built in brms.

## formula ##

```{r}
formula = bf(count ~ 1 + condition +
               (1 + condition | subj))
```

## check the priors available ##

```{r}
get_prior(formula,
          data = d1, family = gaussian())
```

## set priors ##

```{r}
priors <- c(
  set_prior("normal(14, 3)", class = "Intercept"),
  set_prior("normal(0, 1)", class = "b"),
  set_prior("normal(0, 1)", class = "sd"),
  set_prior("normal(0, 1)", class = "sigma"),
  set_prior("lkj(2)", class = "cor")
)
```

change the condition factor coding to make the labels simpler in the model output

```{r}
d1 <- d1 %>% 
  mutate(condition = factor(condition, 
                            levels = c("none", "left", "right")))
head(d1)
str(d1)
```

## fit the model ##

```{r}
t1 <- Sys.time()

plan(multicore)
fit <-
  brm(data = d1,
      family = gaussian(),
      formula = formula,
      prior = priors,
      iter = 2000, warmup = 1000, chains = 4,
      cores = 8, 
      save_pars = save_pars(all=TRUE),
      seed = 1)

t2 <- Sys.time()

t2 - t1

# Time difference of 25s macbook
# iMac?
```

let's take a look 

```{r}
# chains
plot(fit)
# summary
print(fit)
# fixed effects
fixef(fit)

# save initial fit
saveRDS(fit, "models/sims/fit_d1.rds") ## fit based on values derived from model b1
```

## section 4 - update the fit ##

update the fit and check the time taken (to see the time difference when 
compared to not using the 'update' function).

```{r}
set.seed(2)

# create new data
d1u <- add_random(subj = subj_n, rep = rep_n) %>%
  # add and recode categorical variables
  add_within("subj", condition = c("none", "left", "right")) %>%
  add_contrast("condition", "treatment", add_cols = TRUE, 
               colnames = c("left", "right")) %>%
  # add random effects 
  add_ranef("subj", u0s = u0s_sd, u1s = u1s_sd, u2s = u2s_sd,
            .cors = cors) %>%
  add_ranef(sigma = sigma_sd) %>%
  # calculate dv
  mutate(count = b0 + u0s + (b1 + u1s) * left + (b2 + u2s) * right + sigma) %>% 
  mutate(condition = factor(condition, 
                            levels = c("none", "left", "right")))
head(d1u)

# update the fit and supply new data

t1 <- Sys.time()

plan(multicore)
updated_fit <-
  update(fit,
         newdata = d1u,
         seed = 2)

t2 <- Sys.time()

t2 - t1

# Time difference of 4s macbook
```

let's take a look 

```{r}
# chains
plot(updated_fit)
# summary
print(updated_fit)
# fixed effects
fixef(updated_fit)

# save the updated fit
saveRDS(updated_fit, "models/sims/updated_fit_d1u.rds")
```

## section 5 - build a simulation function ##

```{r}
sim1 <- function(subj_n = 12, rep_n = 8,  # these can be changed when calling the function
                b0 = 14.99, b1 = -1.71, b2 = -0.16,        # fixed effects
                u0s_sd = 1.14, u1s_sd = 1.39, u2s_sd = 0.4,   # varying effects
                cors = c(-0.52, -0.11,
                         0.34),   # cor
                sigma_sd = 1.26,           # error term
                ... # helps the function work with pmap() below
                ) {

  # set up data structure
  data <- add_random(subj = subj_n, rep = rep_n) %>%
  # add and recode categorical variables
  add_within("subj", condition = c("none", "left", "right")) %>%
  add_contrast("condition", "treatment", add_cols = TRUE, 
               colnames = c("left", "right")) %>%
  # add random effects 
  add_ranef("subj", u0s = u0s_sd, u1s = u1s_sd, u2s = u2s_sd,
            .cors = cors) %>%
  add_ranef(sigma = sigma_sd) %>%
  # calculate dv
  mutate(count = b0 + u0s + (b1 + u1s) * left + (b2 + u2s) * right + sigma) %>% 
  mutate(condition = factor(condition, 
                            levels = c("none", "left", "right")))

  # glimpse(data) # only use this when testing the code
}
```

test it

```{r}
# sim1(subj_n = 25, rep_n = 10)
```

# section 6 - run the full simulation ##

before running the full thing, I would always first check the pipeline with 2 reps.

## now run it with 1000 reps per variation ##

```{r}
plan(multicore)
x <- crossing(
  repx = 1:1000, # number of replicates
  subj_n = c(8, 12, 16), # range of subject N
  b1 = c(-0.5, -1, -1.71) # effect of the key interaction
) %>%
  mutate(d = pmap(., sim1)) %>%
  mutate(params = map2(d, repx, ~update(fit, newdata = .x, seed = .y) %>% # if you left the code here, then it would store the models and data
                     fixef() %>% 
                     data.frame() %>% 
                     rownames_to_column("parameter"))) %>% 
  select(-d) # adding this line in removes the data from the stored tibble 'x'. 
```

take a look

```{r}
head(x)
```

select parameters of interest to summarise and visualise

```{r}
parameters <-
  x %>% 
  unnest(params)
head(parameters)
```

save out parameters

```{r}
# save the parameters
write_csv(parameters, "data/sims/sim_p1.csv") #  design based on pilot data
```

alternatively, read in saved parameters, if already computed and saved.

```{r}
## read in
# parameters <- read_csv("data/sims/sim_p1.csv") 
# head(parameters)
```

rename parameters and create factors

```{r}
bmx_params <- parameters %>%
  filter(parameter != "Intercept") %>%
  mutate(repx = factor(repx),
         subj_n = factor(subj_n),
         b1 = factor(b1),
         parameter = if_else(parameter == "conditionleft", "left", "right"),
         parameter = factor(parameter, 
                            levels = c("left", "right")))
head(bmx_params)
str(bmx_params)
```

calculate average values

```{r}
bmx_params_qi <- bmx_params %>%
  filter(parameter == "left") %>% 
  group_by(subj_n, b1, parameter) %>% 
  median_qi(Estimate)
head(bmx_params_qi)
```

[note - I think the pull towards zero by the prior might be too strong here. Re-do
sims with a looser prior so that the target values hit the right spot more often]

and plot

```{r}
p_bm_fixed <- ggplot(bmx_params, aes(x = Estimate, y = fct_rev(parameter), 
                                   fill=parameter)) +  
  geom_vline(xintercept = 0, color = "grey", alpha = 5/10) +
  stat_halfeye() +
  labs(title = "Avg. simulated coefficient plot for fixed effects (predictors)",
       x = NULL, y = NULL) +
  theme_bw() +
  scale_fill_brewer(palette = "Dark2") +
  theme(panel.grid   = element_blank(),
        panel.grid.major.y = element_line(color = alpha("firebrick4", 1/2), linetype = 3),
        axis.text.y  = element_text(hjust = 0),
        axis.ticks.y = element_blank(),
        legend.position = "none") +
  coord_cartesian(xlim =c(-2, 2)) +
  scale_x_continuous(breaks=seq(-2,2,0.5)) +
  facet_grid(fct_rev(subj_n)~b1)
p_bm_fixed

ggsave ("figures/sims/bm_p1_fixef.jpeg",
        width = 9, height = 7)
```

calculate power i.e., % Q97.5 < 0

```{r}
bm_power <- bmx_params %>% 
  filter(parameter == "left") %>%
  group_by(subj_n, b1) %>% # here we would group_by stuff that we varied in the sims
  mutate(check = ifelse(Q97.5 < 0, 1, 0)) %>% 
  summarise(power = mean(check))
bm_power
```

plot power

```{r}
p_bm_power <- ggplot(bm_power, aes(x = fct_rev(b1), y = subj_n, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.3f", power)), color = "white", size = 10) +
  scale_fill_viridis_c(limits = c(0, 1))
p_bm_power

ggsave ("figures/sims/bm_p1_power.jpeg")
```

plot parameters and include power as a text label

wrangle

```{r}
bm_plot_params <- bmx_params %>%
  filter(parameter == "left") %>%
  mutate(above_zero = if_else(Q97.5 > 0, "yes", "no"), 
         above_zero = factor(above_zero, levels = c("no", "yes"))) %>% 
  inner_join(bm_power, by = c("subj_n", "b1")) %>% 
  mutate(power = round(power * 100, 2)) 
head(bm_plot_params)
```

plot

```{r}
p_bm_params <- bm_plot_params %>%
  ggplot(aes(x = repx, y = Estimate, ymin = Q2.5, ymax = Q97.5)) +
  geom_pointrange(fatten = 1/2, aes(colour=above_zero)) +
  geom_hline(yintercept = 0, colour = "red") +
  # geom_hline(aes(yintercept = 0.35), colour = "blue") + # this would add at the target effect size
  scale_colour_manual(values=c("darkgrey", "black")) +
  geom_text(aes(x=700, y=2,
                label = sprintf("%.1f%s", power, "% power")), 
            color = "darkgrey", size = 4) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "sim # (i.e., simulation index)",
       y = expression(beta("b1"))) +
  scale_x_discrete(breaks = NULL) +
  facet_grid(fct_rev(subj_n)~fct_rev(b1))
p_bm_params

ggsave ("figures/sims/bm_p1_parameters.jpeg",
        width = 6, height = 6)
```

save out some files

```{r}
write_csv(bmx_params, "data/sims/bmx_p1_params.csv")
write_csv(bmx_params_qi, "data/sims/bmx_p1_param_qi.csv")
```