---
title: "wrangle"
author: "Rich"
date: "2024-01-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file wrangles data for Exp2a of Holmes et al., 2023.

## load the libraries that we will be using ## 

## install ##

```{r install-pkg}
# install.packages(c("tidyverse", "RColorBrewer", "patchwork"))
```

take a snapshot of loaded packages and update the lock.file using renv

```{r snapshot-renv}
# take a snapshot and update the lock.file
# renv::snapshot() # this is only necessary when new packages or installed or packages are updated.
```

## load ##

```{r load-pkg}
pkg <- c("tidyverse", "RColorBrewer", "patchwork")

lapply(pkg, library, character.only = TRUE)
```

## plot settings ##

```{r}
## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

theme settings for ggplot

```{r, eval = F}
theme_set(
  theme_bw() +
    theme(text = element_text(size = 18), 
          title = element_text(size = 18),
          legend.position = "bottom")
)
```

## section 1 - read in the raw data ##

```{r}
raw <- read_delim("data/P10_E8_pegboard.txt", col_names = FALSE)
head(raw)
glimpse(raw)
summary(raw)
```

## section 2 - wrangle into long format ##

```{r}
data <- raw %>% 
  add_column(pid = 1:12, .before = "X1") %>% 
  pivot_longer(-pid,
               names_to = "trial_id",
               values_to = "count") %>%
  mutate(condition = rep(c("left", "right", "none"), each = 8, times = 12),
         trial = rep(1:8, times = 3*12))
head(data)
glimpse(data)
summary(data)
```

remove unnecessary columns, re-order columns and create a factor for condition

```{r}
data <- data %>% 
  select(pid, trial, condition, count) %>%
  mutate(condition = factor(condition,
                            levels = c("left", "none", "right")))
head(data)
```

data check

```{r}
data.check <- data %>% 
  distinct(trial, condition)
data.check
```

pid check

```{r}
pid.check <- data %>% 
  group_by(pid) %>% 
  tally()
pid.check
```

## section 3 - plot the raw data ##

create summary data 

at the pid level

```{r}
summary_pid <- data %>% 
  group_by(pid, condition) %>% 
  summarise(mean = mean(count),
            sd = sd(count),
            n = length(unique(trial)), # n here is the total subjs
            sem = (sd/sqrt(length(unique(trial)))),
            ci = sem*1.96)
summary_pid
```

at the group level

```{r}
summary <- data %>% 
  group_by(condition) %>% 
  summarise(mean = mean(count),
            sd = sd(count),
            n = length(unique(pid)), # n here is the total subjs
            sem = (sd/sqrt(length(unique(pid)))),
            ci = sem*1.96)
summary
```

violin plots

```{r}
p3.1 <- ggplot(summary_pid, aes(x=condition, y = mean, 
                                fill=condition, colour = condition)) +
   geom_jitter(position=position_jitterdodge(dodge.width =1), 
               alpha = 1, colour = "darkgrey") +
   geom_line(aes(group=pid), 
             colour = "darkgrey") +
   geom_violin(alpha = 0.7) +
   geom_point(data = summary, 
             aes(y = mean), size = 3, colour="black") +
   geom_line(data=summary, 
             aes(group=1), colour = "black") +
   geom_errorbar(data = summary,
                aes(y = mean, ymin = mean-sem, ymax = mean +sem),
                width=.2, colour = "black") +
   scale_fill_brewer(palette = "Dark2") +
   scale_colour_brewer(palette = "Dark2") +
   ggtitle("pegboard count by stimulation condition")
p3.1

ggsave("figures/violin.jpeg",
       width = 8, height = 6)
```

not perfect, but does the job for now.

what about difference scores?

## calculate differences score compared to no stimulation ##

at the pid level

```{r}
diff_pid <- summary_pid %>%
  select(pid, condition, mean) %>% 
  pivot_wider(id_cols = c(pid),
              names_from = "condition",
              values_from = "mean") %>% 
  mutate(left_none = left - none,
         right_none = right - none,
         left_right = left - right) %>% 
  select(pid, left_none, right_none, left_right) %>% 
  pivot_longer(cols = -pid,
               names_to = "contrasts",
               values_to = "diff")
head(diff_pid)
```

at the group level

```{r}
diff_group <- diff_pid %>%
  group_by(contrasts) %>% 
  summarise(mean_diff = mean(diff),
            sd = sd(diff),
            sem = (sd/sqrt(10)),
            ci = sem*1.96,
            dz = mean_diff/sd) 
diff_group
```

plot

violin

diff in original units

```{r}
p3.2 <- ggplot(diff_pid, aes(x=contrasts, y = diff)) +
   geom_jitter(width = 0.1, alpha = 1, colour = "darkgrey") +
   geom_violin(alpha = 0.7) +
   geom_point(data = diff_group, 
             aes(y = mean_diff), size = 3, position=pd2, colour="black") +
   geom_errorbar(data = diff_group,
                aes(y = mean_diff, ymin = mean_diff-sem, ymax = mean_diff +sem),
                width=.2, position=pd2, colour = "black") +
   ggtitle("difference score between stimulation conditions") 
p3.2

ggsave("figures/diffs.jpeg",
       width = 8, height = 6)
```

## save out some processed data ##

```{r}
write_csv(data, "data/data.csv")
```

